---
title: "Week 02: CGM QC Features"
author: "Shubho Banerjee"
date: "2026-07-21"
toc: true
toc-location: left
format: html
editor: visual
---

# Introduction

The goal of this notebook is to clean Dexcom G6 continuous glucose monitoring (CGM) data and create summaries at both the participant level and participant-day level. These results will be used later to study how sleep and activity relate to glucose patterns.

I first tested the workflow using one participant and then applied the final pipeline to five CGM participants from the AI-READI mini dataset.

For this assignment, I used all valid CGM days available (up to about 10 days). The AGP-style figure shown here is for learning and visualization purposes only. It should not be considered a clinical 14-day AGP report.

```{r}
library(jsonlite)   # reads the raw Dexcom JSON files
library(dplyr)      # table cleaning/summarizing (filter, mutate, summarise, etc.)
library(lubridate)  # date/time parsing and timezone conversion
library(ggplot2)    # figures
library(scales)     # axis formatting helpers for ggplot2
```

# 1. Identify CGM Participants

Dexcom files are stored in one folder per participant. `list.dirs()` lists every participant sub-folder; `recursive = FALSE` keeps it to one level deep. The first five folders were selected to build and test the pipeline before scaling up.

```{r}
dex_path <- "data/mini/v3-mini/dataset/wearable_blood_glucose/continuous_glucose_monitoring/dexcom_g6"

participants <- list.dirs(
  dex_path,
  recursive = FALSE
)

participants_5 <- participants[1:5]
participants_5
```

# 2. Create CGM Cleaning Function

Each Dexcom JSON file has a header (participant ID, timezone) and a body (the glucose readings). This function turns one raw file into one clean table. `flatten = TRUE` collapses nested JSON fields into single dot-named columns (e.g. `blood_glucose.value`). The `if` block is a safety net that only runs if `dex$body` didn't parse into a clean table already. It forces each reading into its own row and stacks them.

`filter(event_type == "EGV")` keeps only actual glucose readings (EGV = Estimated Glucose Value), excluding calibration/alert events. `transmute()` (rather than `mutate()`) builds a fresh table with only these named columns, dropping every other raw JSON field. `as.numeric(blood_glucose.value)` converts text to numbers; anything non-numeric silently becomes `NA`. This is the source of the missing glucose values found later.

`tz_choice` maps the device's raw timezone label to a proper IANA timezone name. IANA names (e.g. `America/Los_Angeles`) carry real daylight-saving rules for that region, so R can correctly switch between standard/daylight time automatically, as a raw label like "pst" alone can't do that. `TRUE ~ "UTC"` is the fallback if nothing matches.

The final `mutate()` (not `transmute()`, since we want to keep everything built so far) parses the raw timestamp as UTC, converts it to local time, and derives local date/hour and a daylight-saving flag.

```{r}
clean_cgm_file <- function(file_path){
  
  dex <- fromJSON(
    file_path,
    flatten = TRUE
  )
  
  # Extract Dexcom CGM event table
  cgm_egv <- dex$body
  
  if (is.list(cgm_egv) && !"event_type" %in% names(cgm_egv)) {
    
    cgm_egv <- do.call(
      rbind,
      lapply(cgm_egv, as.data.frame)
    )
    
  }
  
  cgm_egv <- as.data.frame(cgm_egv)
  
  cgm_clean <- cgm_egv %>%
    filter(event_type == "EGV") %>%
    transmute(
      person_id = dex$header$patient_id,
      event_type,
      time_utc = effective_time_frame.time_interval.start_date_time,
      timezone_abbreviation = dex$header$timezone,
      glucose = as.numeric(blood_glucose.value),
      glucose_unit = blood_glucose.unit
    )
  
  tz_choice <- case_when(
    cgm_clean$timezone_abbreviation[1] %in% c("pst","pdt") ~ "America/Los_Angeles",
    cgm_clean$timezone_abbreviation[1] %in% c("mst","mdt") ~ "America/Denver",
    cgm_clean$timezone_abbreviation[1] %in% c("cst","cdt") ~ "America/Chicago",
    cgm_clean$timezone_abbreviation[1] %in% c("est","edt") ~ "America/New_York",
    TRUE ~ "UTC"
  )
  
  cgm_clean %>%
    mutate(
      time_utc = ymd_hms(time_utc, tz="UTC"),
      time_local = with_tz(time_utc, tz_choice),
      date_local = as.Date(time_local),
      hour_local = hour(time_local),
      is_dst = dst(time_local)
    )
  
}
```

# 3. Apply Cleaning Pipeline to Five Participants

Runs `clean_cgm_file()` once per participant folder (returning a list of 5 tables), then `bind_rows()` stacks them into one combined table.

```{r}
cgm_all <- lapply(
  participants_5,
  function(folder){
    
    json_file <- list.files(
      folder,
      pattern = "\\.json$",
      full.names = TRUE
    )
    
    clean_cgm_file(json_file[1])
    
  }
)

cgm_all <- bind_rows(cgm_all)

head(cgm_all)

table(cgm_all$person_id)
```

`table()` shows reading counts per participant: roughly 2,800–2,900 each, consistent with \~10 days at 288 readings/day.

The cleaned dataset contains CGM readings from five AI-READI participants.

# 4. Quality Control Checks

## Missing timestamps and glucose values

```{r}
sum(is.na(cgm_all$time_utc))

sum(is.na(cgm_all$glucose))
```

Zero missing timestamps. Six glucose values that couldn't be converted to numbers, treated as missing.

## Duplicate timestamps

A duplicate = same participant ID + same UTC timestamp. Groups by participant + timestamp, counts rows per group, keeps only groups with more than one, with none found here.

```{r}
duplicate_check <- cgm_all %>%
  group_by(person_id,time_utc) %>%
  summarise(
    n=n(),
    .groups="drop"
  ) %>%
  filter(n>1)

duplicate_check
```

General-purpose duplicate handler, built in case duplicates existed: keeps a row if it's not duplicated, or if all its duplicate copies agree on glucose value (`n_distinct(glucose)==1`); disagreeing duplicates get dropped; agreeing duplicates get collapsed to one copy via `slice(1)`. No effect here since no duplicates were found.

```{r}
cgm_final <- cgm_all %>%
  group_by(person_id,time_utc) %>%
  mutate(
    duplicate_n=n(),
    duplicate_glucose_values=n_distinct(glucose)
  ) %>%
  filter(
    duplicate_n==1 |
      duplicate_glucose_values==1
  ) %>%
  slice(1) %>%
  ungroup()
```

## Glucose units

```{r}
unique(cgm_all$glucose_unit)
```

Only "mg/dL" appears.

## Unusual glucose values

Glucose below 40 or above 400 mg/dL is treated as an implausible/erroneous reading for this exercise (a data-processing rule, not a clinical threshold).

```{r}
cgm_all <- cgm_all %>%
  mutate(
    unusual_glucose = glucose < 40 | glucose > 400
  )

table(cgm_all$unusual_glucose)
```

None flagged as unusual here. This next step drops unusual (and, as a side effect, missing) glucose rows so `cgm_final` is now the clean, one-reading-per-row dataset.

```{r}
cgm_final <- cgm_all %>%
  filter(!unusual_glucose)
```

# 5. Determine Valid CGM Days

A CGM day is only trustworthy if the sensor captured enough of it. A "valid day" needs at least 202 of the expected 288 five-minute readings (\~70% coverage), which is the same completeness standard used in clinical CGM interpretation.

```{r}
daily_qc <- cgm_final %>%
  group_by(person_id,date_local) %>%
  summarise(
    n_readings=n(),
    percent_complete=n()/288*100,
    valid_day=n_readings>=202,
    .groups="drop"
  )

daily_qc
```

`inner_join()` keeps only readings whose (participant, date) also appears in `valid_dates`. Readings from invalid days are automatically dropped. This enforces "only valid-day data from here forward."

```{r}
valid_dates <- daily_qc %>%
  filter(valid_day)

cgm_valid <- cgm_final %>%
  inner_join(
    valid_dates,
    by=c("person_id","date_local")
  )

head(cgm_valid)
```

# 6. Create Summary Tables

## Table 1: CGM Quality Control Summary

Shows how much data existed, and how much survived cleaning, per participant. This describes data quality only and it says nothing about whether someone's glucose control was good or bad. `left_join()` attaches each participant's valid-day count (calculated separately from `daily_qc`) onto the QC summary, keeping every row from the summary table.

```{r}
qc_summary <- cgm_all %>%
  group_by(person_id) %>%
  summarise(
    raw_EGV_readings=n(),
    missing_timestamp=sum(is.na(time_utc)),
    missing_glucose=sum(is.na(glucose)),
    duplicate_timestamps=sum(duplicated(time_utc)),
    unusual_values=sum(unusual_glucose),
    cleaned_readings=sum(!unusual_glucose),
    .groups="drop"
  ) %>%
  left_join(
    daily_qc %>%
      group_by(person_id) %>%
      summarise(
        valid_days=sum(valid_day),
        .groups="drop"
      ),
    by="person_id"
  )
```

------------------------------------------------------------------------

## Table 2: Participant-Level CGM Features

One row per participant, summarizing their overall glucose pattern across all valid days. `cv` (coefficient of variation) expresses glucose variability as a percentage of the mean, so variability can be compared across people regardless of average glucose level. Each `pct_*` line takes the average of a TRUE/FALSE condition (proportion of readings meeting it), converted to a percentage since CGM readings occur \~every 5 minutes, % of readings approximates % of time. 70–180 mg/dL is the standard target range; above-250 counts are also included in above-180; below-54 counts are also included in below-70 intentional overlap.

```{r}
participant_summary <- cgm_valid %>%
  group_by(person_id) %>%
  summarise(

    cleaned_readings = n(),

    valid_days = n_distinct(date_local),

    mean_glucose = mean(glucose),
    median_glucose = median(glucose),
    sd_glucose = sd(glucose),

    cv = sd_glucose / mean_glucose * 100,

    pct_70_180 = mean(glucose >= 70 & glucose <= 180, na.rm=TRUE) * 100,

    pct_above_180 = mean(glucose > 180, na.rm=TRUE) * 100,

    pct_above_250 = mean(glucose > 250, na.rm=TRUE) * 100,

    pct_below_70 = mean(glucose < 70, na.rm=TRUE) * 100,

    pct_below_54 = mean(glucose < 54, na.rm=TRUE) * 100,

    .groups="drop"

  )

participant_summary
```

This table summarizes overall glucose patterns per participant. It describes levels and variability, not causes.

------------------------------------------------------------------------

## Table 3: Participant-Day CGM Features

One row per participant, per valid local day. The same features as Table 2, but broken down day by day, plus a daytime/nighttime split. `glucose[hour_local >= 6]` subsets glucose values to daytime hours (6:00 AM–11:59 PM) before averaging; the mirrored condition (`< 6`) covers nighttime (12:00 AM–5:59 AM). This is a clock-time split, not actual sleep/wake timing. It'll be replaced once Garmin sleep data is available.

```{r}
participant_day_summary <- cgm_valid %>%
  group_by(person_id,date_local) %>%
  summarise(

    cleaned_readings = n(),

    percent_complete = n()/288*100,

    mean_glucose = mean(glucose),

    sd_glucose = sd(glucose),

    pct_70_180 = mean(glucose >=70 & glucose <=180)*100,

    pct_above_180 = mean(glucose >180)*100,

    pct_below_70 = mean(glucose <70)*100,


    daytime_readings = sum(hour_local >= 6),

    nighttime_readings = sum(hour_local < 6),


    daytime_mean = mean(glucose[hour_local >= 6]),

    nighttime_mean = mean(glucose[hour_local < 6]),


    daytime_sd = sd(glucose[hour_local >= 6]),

    nighttime_sd = sd(glucose[hour_local < 6]),


    .groups="drop"

  )

participant_day_summary
```

This table is what will eventually be joined to sleep/activity data, day by day.

------------------------------------------------------------------------

# 7. Figures

## Figure 1: Full CGM Trace

One participant's entire glucose trace across their monitoring period. Participant selected: the first one with valid data (a simple, disclosed rule).

```{r}
example_participant <- unique(cgm_valid$person_id)[1]

full_trace <- cgm_valid %>%
  filter(person_id == example_participant)

ggplot(
  full_trace,
  aes(x=time_local,y=glucose)
) +
  geom_line() +
  labs(
    title=paste(
      "Full CGM Trace:",
      example_participant
    ),
    x="Local Time",
    y="Glucose (mg/dL)"
  )
```

Shows glucose rising and falling over time, it doesn't explain why since meals, exercise, medication, and sleep aren't included.

------------------------------------------------------------------------

## Figure 2: Example Valid Local-Day Traces

Two valid days overlaid on one chart to show day-to-day variability. `slice(1:2)` picks the first two valid days found (disclosed selection rule). `semi_join()` filters `cgm_valid` down to just those two days' readings, without attaching extra columns.

```{r}
example_days <- cgm_valid %>%
  group_by(person_id,date_local) %>%
  summarise(
    readings=n(),
    .groups="drop"
  ) %>%
  filter(readings>=202) %>%
  slice(1:2)

two_days <- cgm_valid %>%
  semi_join(
    example_days,
    by=c("person_id","date_local")
  )

ggplot(
  two_days,
  aes(
    x=time_local,
    y=glucose,
    color=factor(date_local)
  )
) +
  geom_line() +
  labs(
    title="CGM Traces for Two Valid Participant-Days",
    x="Local Time",
    y="Glucose (mg/dL)"
  )
```

`color=factor(date_local)` colors each day separately so the two traces are visually distinguishable. Shows that glucose patterns can differ day to day and doesn't explain why.

------------------------------------------------------------------------

## Figure 3: AGP-Style Plot

AGP = Ambulatory Glucose Profile, a standard CGM visualization that overlays all valid days by time-of-day (not calendar date) to show a "typical day" pattern. `time_of_day` = minutes since midnight, letting readings from different days at the same clock time be grouped together. For each time-of-day, calculates the median glucose plus the 25th–75th percentile spread across days.

```{r}
agp_participant <- unique(cgm_valid$person_id)[1]

agp <- cgm_valid %>%
  filter(person_id == agp_participant) %>%
  mutate(
    time_of_day = hour_local*60 + minute(time_local)
  ) %>%
  group_by(time_of_day) %>%
  summarise(

    median_glucose = median(glucose),

    lower = quantile(glucose,0.25),

    upper = quantile(glucose,0.75),

    .groups="drop"

  )
```

Ribbon = day-to-day spread (25th–75th percentile); line = median daily pattern; horizontal lines at 70/180 mg/dL = target range boundaries.

```{r}
ggplot(
  agp,
  aes(x=time_of_day)
) +

  geom_ribbon(
    aes(
      ymin=lower,
      ymax=upper
    ),
    alpha=0.3
  ) +

  geom_line(
    aes(y=median_glucose)
  ) +

  geom_hline(
    yintercept=70
  ) +

  geom_hline(
    yintercept=180
  ) +

  labs(
    title=paste(
      "AGP-Style CGM Profile:",
      agp_participant
    ),
    x="Minutes After Midnight",
    y="Glucose (mg/dL)"
  )
```

This is an educational AGP-style visualization, not a clinical AGP. Clinical AGP interpretation typically requires 14 days of ≥70%-complete data, and this dataset has up to \~10 days.

------------------------------------------------------------------------

# 8. Week 3 Preparation

## Which participants have at least one valid local day?

```{r}
unique(cgm_valid$person_id)
```

All five participants had at least one valid local CGM day.

------------------------------------------------------------------------

## Which local dates are valid for each participant?

```{r}
valid_dates
```

Shows exactly which participant-date combinations passed the coverage threshold, and by omission, where the gaps are.

------------------------------------------------------------------------

## Which timestamp will be used to align CGM with sleep/activity intervals?

`time_local` will be used, since sleep and activity data are also recorded in local time.

------------------------------------------------------------------------

## Does the participant-day table have exactly one row per person_id + date_local combination?

```{r}
participant_day_summary %>%
  count(person_id,date_local) %>%
  filter(n>1)
```

Empty result confirms no duplicate participant-day rows and the table is at the correct grain for merging with other data later.

------------------------------------------------------------------------

## Is the participant-day table ready to join with sleep/activity data?

Yes—one row per participant-day, no duplicate keys, and glucose features already summarized. Future work will scale this pipeline to more participants and merge in wearable sleep/activity data.

------------------------------------------------------------------------

# Week 3 Presentation Summary

For the Week 3 update:

- CGM data were cleaned and checked for missing values, duplicate timestamps, incorrect units, and unusual glucose values.

- CGM visualizations were created, including full traces and an AGP-style plot.

- Participant-level and participant-day glucose features were calculated.

- A challenge was processing Dexcom data stored as nested JSON files and converting timestamps correctly.

- The next step is to study how sleep and activity patterns relate to individual glucose changes.
